### 2021-06-29 rad-icfp-2021/correctness.tex
minors
writing appendix
changing outline
simple writing phase

\begin{theorem}
    For every $\Gamma \vdash$ !e: \reals!, we have 
    $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
\end{theorem}
where $Id_\Gamma$ is the identity function, 
$\sem{-}$ an appropriate denotational semantics of smooth functions,
and $last(a)$ is the hot-vecteur encoding $(0,\ldots,0,a)$.

\subsection{Denotational semantics}

\begin{itemize}
    \item for Syn
    \item for target
    \item for UNF
\end{itemize}

\MH{find somewhere to put the operational semantics}
\input{figures/op_semantics_target}

\subsection{Correctness}

In order to prove correctness, we need some key intermediate lemmas.

\begin{lemma}
    $\sem{\pi_1\invUNF{\UNF{e}}}=\sem{e}$.
\end{lemma}

\begin{lemma}
    $\sem{\pi_2\invUNF{e}}=T^*\sem{\pi_1\invUNF{e}}$
\end{lemma}

From this, we deduce the main theorem

\begin{theorem}
    \MH{slight generalisation of the corrolary for compositionality reasons}
\end{theorem}

The main theorem specializes to our case of interest

\begin{corollary}
    For every $\Gamma \vdash$ !e: \reals!, we have 
    $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
\end{corollary}

>>>>

 \begin{theorem}
    TODO: rephrase following Section~\ref{sec:simplediff}.
     For every $\Gamma \vdash$ !e: \reals!, we have 
     $\sem{\pi_2 \invUNF{\UNF{e}}(Id_\Gamma)(last(1:\reals))}= \nabla_{\Gamma}\sem{e}$.
 \end{theorem}
 where $Id_\Gamma$ is the identity function, 
 $\sem{-}$ an appropriate denotational semantics of smooth functions,
 and $last(a)$ is the hot-vecteur encoding $(0,\ldots,0,a)$.

\subsection{Denotational semantics for the source and target languages}

\begin{itemize}
    \item syntactic (multi?)category category
    \item semantics in Manifolds or Eucl: TODO: need be clearer arrays of fixed size in syntax, should reflect the semantics
    \item semantics for the array part
\end{itemize}

 \subsection{UNF semantics} % (fold)
 \label{sub:UNF semantics}

\begin{itemize}
    \item TODO: what do I do here?
\end{itemize}

 \subsection{Correctness transformations} % (fold)
 \label{sub:Correctness transformations}

 \subsection{Correctness optimisations} % (fold)
 \label{sub:Correctness optimisations}

### 2021-06-29 rad-icfp-2021/purerad.tex
simple writing phase

We consider a standard call-by-value language. 
It is a first-order language with array constructs and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are in Figure~\ref{fig:source_grammar}.

@@

The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
but it is not difficult to add further operations. For scalar operations, we assume given a set of operations. 
!op1! and !op2! denote respectively unary and binary smooth operations on reals. 
We use infix notation for binary operators.
Typical examples include !cos, exp, +, *!. 
For the sake of simplicity in the development, the function in the array operations !map, map2, foldl! are quite restricted with and free variables. 
We also restrict to arrays of reals.
The general case is also useful and presented in Section~\ref{sec:generalisation}.

@@

The last language we present, which we call the target language, is an extension to the source language. 
The types and terms are presented in Figure~\ref{fig:target_grammar}.
The language contains lambda abstractions for dealing with continuations. 
We are not interested here with partial applications and our language is pure, so functions take $n$ arguments.
Lambda abstractions and applications are going to be removed during partial evaluation and this notation makes reading slightly easier.

@@

In addition, we require the set of operations to be closed under partial differentiation. 
In more detail, for every unary scalar operation !op1!, 
we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
representing respectively the first and second partial derivative of !op2!.
Similarly, we need to add some functions on arrays. 
!scanl! is a !foldl! that stores all intermediate results. 
!scanr! is similar, but reads the array from right to left, and also stores from right to left.
!reduce! is a !foldl! for which the function is associative. 
This means in practice that it can be computed faster.
We also include a !map3! operator. 
Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can somtimes run faster in practice.
Finally, we add a somewhat strange operator which we call !shift1!. 
It takes an array of size $n$,  shifts all the element of the array by one, 
forgetting the last element, and puts a one in the first place. 
This operator naturally shows up when differentiating !foldl!.

The type system for the extended grammar of target is in Figure~\ref{fig:target_typesystem}.

@@

We now present our transformation from target UNF to target. 
The transformation is presented in Figure~\ref{fig:unf_to_target}.

\input{figures/unf_to_target}

@@

@@

% subsectionPartial evaluation and optimisation (end)

>>>>

We consider a standard call-by-value language. 
It consist in a first-order functional language enriched with arrays and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.

@@

The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
but it is not difficult to add further operations. 
For scalar operations, we assume given a set of operations, including $+$ and $*$. 
!op1! and !op2! denote respectively a unary and a binary smooth operation on reals. 
Typical examples include !cos, exp, +, *!. 
We use infix notation for binary operators.
For the sake of simplicity in the presentation, a bound function in an array operation !map, map2, foldl! is restricted to having no free variables.
For the same reason, we restrict arrays of reals.
The general cases are presented in Section~\ref{sec:generalisation}.

@@

The target language for differentiation is an extension to the source language, 
which is higher-order as needed to carry the continuation used to have a pure reverse-mode transformation, 
and contains more array primitives. In addition, we require the set of operations to be closed under partial differentiation.

The types and terms are presented in Figure~\ref{fig:target_grammar}.
We are not interested here with partial applications and our language is pure, so functions take $n$ arguments.
Lambda abstractions and applications will be removed during partial evaluation and this notation makes reading slightly easier.

@@

In more detail, for every unary scalar operation !op1!, 
we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
representing respectively the first and second partial derivative of !op2!.
Similarly, some additional operations on arrays are needed. 
!scanl! is a !foldl! that stores all intermediate results. 
!scanr! is similar, but reads the array from right to left, and also stores from right to left.
!reduce! is a !foldl! for which the function is associative. 
This means in practice that it can be computed faster.
We also include a !map3! operator. 
Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can sometimes run faster in practice.
Finally, we add a somewhat strange operator which we call !shift1!. 
It takes an array of size $n$,  shifts all the element of the array by one, 
forgetting the last element, and puts a one in the first place. 
This operator naturally shows up when differentiating !foldl!.

The type system for the extended grammar of target is presented in Figure~\ref{fig:target_typesystem}.

@@

@@

In Figure~\ref{fig:direct_diff_macro} we present the direct transformation for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
This transformation may be hard to read and prove correct. 
In Section~\ref{sec:unf} we decompose this transformation into several easier 
and more intuitive steps via a new intermediate representation which we call UNF.

TODO: on the examples

\input{figures/direct_macro_diff}

@@

TODO: how to get gradient or subpart of the gradient from the previosu term.

TODO: maybe intuition around here

TODO: on the examples

TODO: partial evaluation to get a tuple (i.e no lambdas)

TODO: use the optimisations to be efficient, just on example?

The full list of optimisations can be found on Figure~\ref{fig:optim}.
